---
title: "Hands-on Exercise 3"
description: |
  A short description of the post.
author:
  - name: Ngah Xin Yan
    url: https://github.com/nxinyan/
date: 08-30-2021
output:
  distill::distill_article:
    self_contained: false
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


## Installing and Loading the R packages

```{r echo=TRUE, eval=TRUE}

packages = c('sf', 'tmap', 'tidyverse')
for (p in packages){
  if(!require(p, character.only = T)){
    install.packages(p)
  }
  library(p,character.only = T)
}

```

## Importing Geospatial Data
                
```{r echo=TRUE, eval=TRUE}
mpsz <- st_read(dsn = "data/geospatial", 
                layer = "MP14_SUBZONE_WEB_PL")
```

```{r echo=TRUE, eval=TRUE}
mpsz
```
## Importing Attribute Data

```{r echo=TRUE, eval=TRUE}
popdata <- read_csv("data/aspatial/respopagesextod2011to2020.csv")
```
## Data Preparation

**1. DATA WRANGLING**

The following data wrangling and transformation functions will be used:

- pivot_wider() of tidyr package, and
- mutate(), filter(), group_by() and select() of dplyr package

```{r echo=TRUE, eval=TRUE}
popdata2020 <- popdata %>%
  filter(Time == 2020) %>%
  group_by(PA, SZ, AG) %>%
  summarise(`POP` = sum(`Pop`)) %>%
  ungroup()%>%
  pivot_wider(names_from=AG, 
              values_from=POP) %>%
  mutate(YOUNG = rowSums(.[3:6])
         +rowSums(.[12])) %>%
mutate(`ECONOMY ACTIVE` = rowSums(.[7:11])+
rowSums(.[13:15]))%>%
mutate(`AGED`=rowSums(.[16:21])) %>%
mutate(`TOTAL`=rowSums(.[3:21])) %>%  
mutate(`DEPENDENCY` = (`YOUNG` + `AGED`)
/`ECONOMY ACTIVE`) %>%
  select(`PA`, `SZ`, `YOUNG`, 
       `ECONOMY ACTIVE`, `AGED`, 
       `TOTAL`, `DEPENDENCY`)
```
**2. JOINING THE ATTRIBUTE DATA AND GEOSPATIAL DATA**

```{r echo=TRUE, eval=TRUE}
popdata2020 <- popdata2020 %>%
  mutate_at(.vars = vars(PA, SZ), 
          .funs = funs(toupper)) %>%
  filter(`ECONOMY ACTIVE` > 0)
```
- Join the geographical data and attribute table using planning subzone name

```{r echo=TRUE, eval=TRUE}
mpsz_pop2020 <- left_join(mpsz, popdata2020,
                          by = c("SUBZONE_N" = "SZ"))
```



## Choropleth Mapping Geospatial Data Using tmap

### Plotting a choropleth map quickly by using qtm()

```{r echo=TRUE, eval=TRUE}
tmap_mode("plot")
qtm(mpsz_pop2020, 
    fill = "DEPENDENCY")
```

### Creating a choropleth map by using tmap’s elements

```{r echo=TRUE, eval=TRUE}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY", 
          style = "quantile", 
          palette = "Blues",
          title = "Dependency ratio") +
  tm_layout(main.title = "Distribution of Dependency Ratio by planning subzone",
            main.title.position = "center",
            main.title.size = 1.2,
            legend.height = 0.45, 
            legend.width = 0.35,
            frame = TRUE) +
  tm_borders(alpha = 0.5) +
  tm_compass(type="8star", size = 2) +
  tm_scale_bar() +
  tm_grid(alpha =0.2) +
  tm_credits("Source: Planning Sub-zone boundary from Urban Redevelopment Authorithy (URA)\n and Population data from Department of Statistics DOS", 
             position = c("left", "bottom"))
```
**1. DRAWING A BASE MAP**

```{r echo=TRUE, eval=TRUE}
tm_shape(mpsz_pop2020) +
  tm_polygons()
```
**2. DRAWING A CHOROPLETH MAP USING TM_POLYGONS()**

```{r echo=TRUE, eval=TRUE}
tm_shape(mpsz_pop2020)+
  tm_polygons("DEPENDENCY")
```

**3. DRAWING A CHOROPLETH MAP USING TM_FILL() AND TM_BORDER()**

```{r echo=TRUE, eval=TRUE}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY")
```

```{r echo=TRUE, eval=TRUE}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY") +
  tm_borders(lwd = 0.1, alpha = 1)
```




### Data classification methods of tmap

**1. PLOTTING CHOROPLETH MAPS WITH BUILT-IN CLASSIFICATION METHODS**

```{r echo=TRUE, eval=TRUE}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          n = 5,
          style = "jenks") +
  tm_borders(alpha = 0.5)
```
- Equal data classification is used in the code below

```{r echo=TRUE, eval=TRUE}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          n = 5,
          style = "equal") +
  tm_borders(alpha = 0.5)
```
- Notice that distribution of quantile data classification method are more evenly distributed then equal data classification method

**2. PLOTTING CHOROPLETH MAP WITH CUSTOME BREAK**

```{r echo=TRUE, eval=TRUE}
summary(mpsz_pop2020$DEPENDENCY)
```                

```{r echo=TRUE, eval=TRUE}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          breaks = c(0, 0.60, 0.70, 0.80, 0.90, 1.00)) +
  tm_borders(alpha = 0.5)
```
***DIY: Using what you had learned, prepare choropleth maps by using different classification methods supported by tmap and compare their differences.***  

**SD Data Classification**

```{r echo=TRUE, eval=TRUE}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          n = 5,
          style = "sd") +
  tm_borders(alpha = 0.5)
```
**Quantile Data Classification**

```{r echo=TRUE, eval=TRUE}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          n = 5,
          style = "quantile") +
  tm_borders(alpha = 0.5)
```

***DIY: Preparing choropleth maps by using similar classification method but with different numbers of classes (i.e. 2, 6, 10, 20). Compare the output maps, what observation can you draw?***  

**Number of classes,n = 2 for quantile and equal data classification**

Quantile data classification

```{r echo=TRUE, eval=TRUE}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          n = 2,
          style = "quantile") +
  tm_borders(alpha = 0.5)
```

Equal data classification

```{r echo=TRUE, eval=TRUE}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          n = 2,
          style = "equal") +
  tm_borders(alpha = 0.5)
```

Observation: Both maps looks like it is more generalised, could not tell more details from it.

**Number of classes,n = 6 for quantile and equal data classification**

Quantile data classification

```{r echo=TRUE, eval=TRUE}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          n = 6,
          style = "quantile") +
  tm_borders(alpha = 0.5)
```

Equal data classification

```{r echo=TRUE, eval=TRUE}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          n = 6,
          style = "equal") +
  tm_borders(alpha = 0.5)
```
Observation: As the number of classes increases, for quantile map, it becomes more evenly distributed, as for equal data maps it stayed the same

### Colour Scheme
**1. USING COLOURBREWER PALETTE**

```{r echo=TRUE, eval=TRUE}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          n = 6,
          style = 'quantile',
          palette = 'Blues') +
  tm_borders(alpha = 0.5)
```

**Reverse colour shading for blue**

```{r echo=TRUE, eval=TRUE}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY", 
          style = "quantile",
          palette = "-Blues") + 
  tm_borders(alpha = 0.5)
```                
                
### Map Layouts
**MAP LEGEND**
                
```{r echo=TRUE, eval=TRUE}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY", 
          style = "jenks", 
          palette = "Blues", 
          legend.hist = TRUE, 
          legend.is.portrait = TRUE,
          legend.hist.z = 0.1) + 
  tm_layout(main.title = "Distribution of Dependency Ration by planning subzone \n(Jenks classification)",
            main.title.position = "center",
            main.title.size = 1,
            legend.height = 0.45, 
            legend.width = 0.35,
            legend.outside = FALSE,
            legend.position = c("right", "bottom"),
            frame = FALSE) +
  tm_borders(alpha = 0.5)
```
**2. MAP STYLE**

```{r echo=TRUE, eval=TRUE}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY", 
          style = "quantile", 
          palette = "-Greens") +
  tm_borders(alpha = 0.5) +
  tmap_style("classic")
```                      
**3. CARTOGRAPHIC FURNITURE**     
                
```{r echo=TRUE, eval=TRUE}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY", 
          style = "quantile", 
          palette = "Blues",
          title = "No. of persons") +
  tm_layout(main.title = "Distribution of Dependency Ratio \nby planning subzone",
            main.title.position = "center",
            main.title.size = 1.2,
            legend.height = 0.45, 
            legend.width = 0.35,
            frame = TRUE) +
  tm_borders(alpha = 0.5) +
  tm_compass(type="8star", size = 2) +
  tm_scale_bar(width = 0.15) +
  tm_grid(lwd = 0.1, alpha = 0.2) +
  tm_credits("Source: Planning Sub-zone boundary from Urban Redevelopment Authorithy (URA)\n and Population data from Department of Statistics DOS", 
             position = c("left", "bottom"))
```
To reset default style

```{r echo=TRUE, eval=TRUE}
tmap_style("cobalt")
```
### Drawing Small Multiple Choropleth Maps
**1. BY ASSIGNING MULTIPLE VALUES TO AT LEAST ONE OF THE AESTHETIC ARGUMENTS**
- showing same map with different factor


```{r echo=TRUE, eval=TRUE}
tm_shape(mpsz_pop2020)+
  tm_fill(c("YOUNG", "AGED"),
          style = "equal", 
          palette = "Blues") +
  tm_layout(legend.position = c("right", "bottom")) +
  tm_borders(alpha = 0.5) +
  tmap_style("white")
```       

```{r echo=TRUE, eval=TRUE}
tm_shape(mpsz_pop2020)+ 
  tm_polygons(c("DEPENDENCY","AGED"),
          style = c("equal", "quantile"), 
          palette = list("Blues","Greens")) +
  tm_layout(legend.position = c("right", "bottom"))
```       

**2. BY DEFINING A GROUP-BY VARIABLE IN TM_FACETS()**

```{r echo=TRUE, eval=TRUE}
tm_shape(mpsz_pop2020) +
  tm_fill("DEPENDENCY",
          style = "quantile",
          palette = "Blues",
          thres.poly = 0) + 
  tm_facets(by="REGION_N", 
            free.coords=TRUE, 
            drop.shapes=TRUE) +
  tm_layout(legend.show = FALSE,
            title.position = c("center", "center"), 
            title.size = 20) +
  tm_borders(alpha = 0.5)
```       
**3. BY CREATING MULTIPLE STAND-ALONE MAPS WITH TMAP_ARRANGE()**

```{r echo=TRUE, eval=TRUE}
youngmap <- tm_shape(mpsz_pop2020)+ 
  tm_polygons("YOUNG", 
              style = "quantile", 
              palette = "Blues")

agedmap <- tm_shape(mpsz_pop2020)+ 
  tm_polygons("AGED", 
              style = "quantile", 
              palette = "Blues")

tmap_arrange(youngmap, agedmap, asp=1, ncol=2)
```  

### Mappping Spatial Object Meeting a Selection Criterion
- a map to show two factor at one time (e.g. central region and dependency)

```{r echo=TRUE, eval=TRUE}
tm_shape(mpsz_pop2020[mpsz_pop2020$REGION_N=="CENTRAL REGION", ])+
  tm_fill("DEPENDENCY", 
          style = "quantile", 
          palette = "Blues", 
          legend.hist = TRUE, 
          legend.is.portrait = TRUE,
          legend.hist.z = 0.1) +
  tm_layout(legend.outside = TRUE,
            legend.height = 0.45, 
            legend.width = 5.0,
            legend.position = c("right", "bottom"),
            frame = FALSE) +
  tm_borders(alpha = 0.5)
```

