---
title: "In-class Exercise 4"
description: |
  A short description of the post.
author:
  - name: Ngah Xin Yan
    url: https://github.com/nxinyan/
date: 09-06-2021
output:
  distill::distill_article:
    self_contained: false
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


# Installing and Loading the R packages

- sf is mainly used for importing geospatial data.
- raster for handling raster-based geospatial data 
- spatstat is used for spatial point analysis


```{r echo=TRUE, eval=TRUE}

packages = c('maptools', 'sf', 'raster','spatstat', 'tmap','tidyverse' )
for (p in packages){
if(!require(p, character.only = T)){
install.packages(p)
}
library(p,character.only = T)
}

```

## Importing Geospatial Data

```{r echo=TRUE, eval=TRUE}
sg_sf <- st_read(dsn = "data/shapefile", 
                 layer="CostalOutline")
mpsz_sf <- st_read(dsn = "data/shapefile", 
                 layer="MP14_SUBZONE_WEB_PL")
```
## Importing Aspatial Data

```{r echo=TRUE, eval=TRUE}
childcare <- read_rds("data/rds/childcare.rds")
CHAS <- read_rds("data/rds/CHAS.rds")
```

### Converting from aspatial to geospatial

```{r echo=TRUE, eval=TRUE}
CHAS_sf <- st_as_sf(CHAS,
                    coords = c("X_COORDINATE",
                               "Y_COORDINATE"),
                    crs=3414)
```

```{r echo=TRUE, eval=TRUE}
childcare$Lat <- as.numeric(childcare$Lat)
childcare$Lng <- as.numeric(childcare$Lng)
childcare_sf <- st_as_sf(childcare,
                         coords = c("Lng",
                                    "Lat"),
                         crs=4326) %>%
  st_transform(crs = 3414)
```

## Geospatial Data Wrangling

### Converting from sf to Spatial Classes

```{r echo=TRUE, eval=TRUE}
childcare <- as_Spatial(childcare_sf)
CHAS <- as_Spatial(CHAS_sf)
mpsz <- as_Spatial(mpsz_sf)
sg <- as_Spatial(sg_sf)
```

### Conerting from Spatial* classes to sp format

```{r echo=TRUE, eval=TRUE}
childcare_sp <- as(childcare, "SpatialPoints")
CHAS_sp <- as(CHAS, "SpatialPoints")
sg_sp <- as(sg, "SpatialPolygons")
```

### Converting from sp to spatstat ppp format

```{r echo=TRUE, eval=TRUE}
childcare_ppp <- as(childcare_sp, "ppp")
CHAS_ppp <- as(CHAS_sp, "ppp")
```

```{r echo=TRUE, eval=TRUE}
tmap_mode('view')
tm_shape(childcare_sf) + 
  tm_dots(alpha=0.4,
          col="blue",
          size=0.05) +
  tm_shape(CHAS_sf) + 
    tm_dots(alpha=0.4,
            col="red",
            size=0.05)
```

```{r echo=TRUE, eval=TRUE}

```

```{r echo=TRUE, eval=TRUE}

```

```{r echo=TRUE, eval=TRUE}

```

```{r echo=TRUE, eval=TRUE}

```

```{r echo=TRUE, eval=TRUE}

```

```{r echo=TRUE, eval=TRUE}

```

```{r echo=TRUE, eval=TRUE}

```

```{r echo=TRUE, eval=TRUE}```
