---
title: "In-class Exercise 3"
description: |
  A short description of the post.
author:
  - name: Ngah Xin Yan
    url: https://github.com/nxinyan/
date: 08-31-2021
output:
  distill::distill_article:
    self_contained: false
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


## Installing and Loading the R packages

```{r echo=TRUE, eval=TRUE}

packages = c('sf', 'tmap', 'tidyverse')
for (p in packages){
  if(!require(p, character.only = T)){
    install.packages(p)
  }
  library(p,character.only = T)
}

```

## Importing Geospatial Data
                
```{r echo=TRUE, eval=TRUE}
mpsz <- st_read(dsn = "data/geospatial", 
                layer = "MP14_SUBZONE_WEB_PL")
```

```{r echo=TRUE, eval=TRUE}
mpsz
```
## Importing Attribute Data

```{r echo=TRUE, eval=TRUE}
popdata <- read_csv("data/aspatial/respopagesextod2011to2020.csv")
```
## Data Preparation

**1. DATA WRANGLING**

The following data wrangling and transformation functions will be used:

- pivot_wider() of tidyr package, and
- mutate(), filter(), group_by() and select() of dplyr package

```{r echo=TRUE, eval=TRUE}
popdata2020 <- popdata %>%
  filter(Time == 2020) %>%
  group_by(PA, SZ, AG) %>%
  summarise(`POP` = sum(`Pop`)) %>%
  ungroup()%>%
  pivot_wider(names_from=AG, 
              values_from=POP) %>%
  mutate(YOUNG = rowSums(.[3:6])
         +rowSums(.[12])) %>%
mutate(`ECONOMY ACTIVE` = rowSums(.[7:11])+
rowSums(.[13:15]))%>%
mutate(`AGED`=rowSums(.[16:21])) %>%
mutate(`TOTAL`=rowSums(.[3:21])) %>%  
mutate(`DEPENDENCY` = (`YOUNG` + `AGED`)
/`ECONOMY ACTIVE`) %>%
  select(`PA`, `SZ`, `YOUNG`, 
       `ECONOMY ACTIVE`, `AGED`, 
       `TOTAL`, `DEPENDENCY`)
```
**2. JOINING THE ATTRIBUTE DATA AND GEOSPATIAL DATA**

```{r echo=TRUE, eval=TRUE}
popdata2020 <- popdata2020 %>%
  mutate_at(.vars = vars(PA, SZ), 
          .funs = funs(toupper)) %>%
  filter(`ECONOMY ACTIVE` > 0)
```
- Join the geographical data and attribute table using planning subzone name

```{r echo=TRUE, eval=TRUE}
mpsz_pop2020 <- left_join(mpsz, popdata2020,
                          by = c("SUBZONE_N" = "SZ"))
```

**Boxplot**

```{r echo=TRUE, eval=TRUE}
ggplot(data=mpsz_pop2020,
       aes(x = "",
           y = AGED)) +
  geom_boxplot()
```       

```{r echo=TRUE, eval=TRUE}
mpszpop2020a <-mpsz_pop2020 %>%
  drop_na()
```
      
**Function to extract a variable as a vector out of an sf data frame**

```{r echo=TRUE, eval=TRUE}
get.var <- function(vname,df) {
v <- df[vname] %>%
st_set_geometry(NULL)
v <- unname(v[,1])
return(v)
}
```  
   
**Percentile Mapping Function**
```{r echo=TRUE, eval=TRUE}
percentmap <- function(vnam,df,legtitle=NA, mtitle="Percentile Map"){
  percent <- c(0,.01,.1,.5,.9,.99,1)
  var <- get.var(vnam,df)
  bperc <- quantile(var,percent)
  tm_shape(mpsz_pop2020) +
  tm_polygons() +
  tm_shape(df) + 
  tm_fill(vnam,
          title = legtitle,
          breaks=bperc,
          palette="Blues",
        labels=c("< 1%", "1% - 10%",
                   "10% - 50%",
                   "50% - 90%",
                   "90% - 99%",
                   "> 99%")) +
  tm_borders() +
  tm_layout(title = mtitle, title.position = c("right","bottom"))
}
``` 

 

```{r echo=TRUE, eval=TRUE}
youngmap <- percentmap("YOUNG", mpszpop2020a)
agedmap <- percentmap("AGED", mpszpop2020a)
```

```{r echo=TRUE, eval=TRUE}
youngmap
agedmap
```       

### Boxbreaks Function

```{r echo=TRUE, eval=TRUE}
boxbreaks <- function(v,mult=1.5) 
  {
  qv <- unname(quantile(v))
  iqr <- qv[4] - qv[2]
  upfence <- qv[4] + mult * iqr
  lofence <- qv[2] - mult * iqr
  # initialize break points vector
  bb <- vector(mode="numeric",length=7)
  # logic for lower and upper fences
  if (lofence < qv[1]) { # no lower outliers
    bb[1] <- lofence
    bb[2] <- floor(qv[1])
  } else {
    bb[2] <- lofence
    bb[1] <- qv[1]
  }
  if (upfence > qv[5]) { # no upper outliers
    bb[7] <- upfence
    bb[6] <- ceiling(qv[5])
  } else {
    bb[6] <- upfence
    bb[7] <- qv[5]
  }
  bb[3:5] <- qv[2:4]
  return(bb)
}

```       

```{r echo=TRUE, eval=TRUE}
get.var <- function(vname,df) {
v <- df[vname] %>% st_set_geometry(NULL)
v <- unname(v[,1])
return(v)
}

```       

```{r echo=TRUE, eval=TRUE}
mpszpop2020a <- mpsz_pop2020 %>%
filter(AGED>=0)
var <- get.var("AGED"
, mpszpop2020a)
boxbreaks(var)


```
### Box Map Function

```{r echo=TRUE, eval=TRUE}
boxmap <- function(vnam, df,
                   legtitle=NA,
                   mtitle="Box Map",
                   mult=1.5){
  var <- get.var(vnam,df)
  bb <- boxbreaks(var)
  tm_shape(mpsz_pop2020) +
  tm_polygons() +
  tm_shape(df) +
    tm_fill(vnam,title=legtitle,
    breaks=bb,
    palette="Blues",
  labels = c("lower outlier",
             "< 25%",
             "25% - 50%",
             "50% - 75%",
             "> 75%",
             "upper outlier")) +
  tm_borders() +
  tm_layout(title = mtitle, 
            title.position = c("right",
                               "bottom"))
}
```       

```{r echo=TRUE, eval=TRUE}
boxmap("ECONOMY ACTIVE", mpszpop2020a)
```       

### Raw Rate Map

```{r echo=TRUE, eval=TRUE}
mpszpop2020a <- mpsz_pop2020 %>%
mutate(`AGED%` = (`AGED`
/`TOTAL`)*100) %>%
filter(`AGED%` >= 0)

```       

```{r echo=TRUE, eval=TRUE}
var <- get.var("AGED%", mpszpop2020a)
boxbreaks(var)
boxmap("AGED%", mpszpop2020a)

```
