---
title: "In-class Exercise 2"
description: |
  In this hands-on exercise, I learn how to handle geospatial data in R by using sf package.
author:
  - name: Ngah Xin Yan
    url: https://github.com/nxinyan/
date: 08-23-2021
output:
  distill::distill_article:
    self_contained: false
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Getting Started

This code chunk performs three tasks:

- create a list of R packages needed.

```{r echo=TRUE, eval=TRUE}

packages <- c('sf', 'tidyverse')
for (p in packages){
  if(!require(p, character.only = T)){
    install.packages(p)
  }
  library(p,character.only = T)
}

```

```{r echo=TRUE, eval=TRUE}
mpsz = st_read(dsn = "data/geospatial", 
                  layer = "MP14_SUBZONE_WEB_PL")
```

```{r echo=TRUE, eval=TRUE}
cyclingpath = st_read(dsn = "data/geospatial", 
                         layer = "CyclingPath")
```

```{r echo=TRUE, eval=TRUE}
preschool = st_read("data/geospatial/pre-schools-location-kml.kml")
```

```{r echo=TRUE, eval=TRUE}
st_geometry(mpsz)
glimpse(mpsz)
head(mpsz, n=5)  
```

```{r echo=TRUE, eval=TRUE}
plot(mpsz)
```

```{r echo=TRUE, eval=TRUE}
plot(st_geometry(mpsz))
```

```{r echo=TRUE, eval=TRUE}
plot(mpsz["PLN_AREA_N"])
```

```{r echo=TRUE, eval=TRUE}
st_crs(mpsz)
```

```{r echo=TRUE, eval=TRUE}
mpsz3414 <- st_set_crs(mpsz, 3414)
```

```{r echo=TRUE, eval=TRUE}
st_crs(mpsz3414)
```

```{r echo=TRUE, eval=TRUE}
preschool3414 <- st_transform(preschool, 
                              crs = 3414)
st_geometry (preschool3414)
```

Import aspatial data as a tibble data frame

```{r echo=TRUE, eval=TRUE}
listings <- read_csv("data/aspatial/listings.csv")
```
Examine Data File

```{r echo=TRUE, eval=TRUE}
list(listings) 
```

```{r echo=TRUE, eval=TRUE}
listings_sf <- st_as_sf(listings, 
                       coords = c("longitude", "latitude"),
                       crs=4326) %>%
  st_transform(crs = 3414)
```

Reveals the data type of each fields

```{r echo=TRUE, eval=TRUE}
glimpse(listings_sf)
```
## Geoprocessing with sf package

### The scenario:

The authority is planning to upgrade the exiting cycling path. To do so, they need to acquire 5 metres of reserved land on the both sides of the current cycling path. You are tasked to determine the extend of the land need to be acquired and their total area.

### The solution:

First, compute the 5-meter buffers around cycling paths

```{r echo=TRUE, eval=TRUE}
buffer_cycling <- st_buffer(cyclingpath, 
                               dist=5, nQuadSegs = 30)
```

Next, calculate the area of the buffers

```{r echo=TRUE, eval=TRUE}
buffer_cycling$AREA <- st_area(buffer_cycling)
```

Lastly, sum() of Base R will be used to derive the total land involved

```{r echo=TRUE, eval=TRUE}
sum(buffer_cycling$AREA)
```

## Point-in-polygon count

### The scenario:

A pre-school service group want to find out the numbers of pre-schools in each Planning Subzone.

### The solution:

Firstly, identify pre-schools located inside each Planning Subzone by using st_intersects(). 
Next, length() of Base R is used to calculate numbers of pre-schools that fall inside each planning subzone.

```{r echo=TRUE, eval=TRUE}
mpsz3414$`PreSch Count`<- lengths(st_intersects(mpsz3414, preschool3414))
```

Check summary of statistics

```{r echo=TRUE, eval=TRUE}
summary(mpsz3414$`PreSch Count`)
```

Lastly, list the planning subzone with the most number of pre-school

```{r echo=TRUE, eval=TRUE}
top_n(mpsz3414, 1, `PreSch Count`)
```
### The solution:

Firstly, derive the area of each planning subzone

```{r echo=TRUE, eval=TRUE}
mpsz3414$Area <- mpsz3414 %>%
  st_area()
```

Next, mutate() of dplyr package is used to compute the density

```{r echo=TRUE, eval=TRUE}
mpsz3414 <- mpsz3414 %>%
  mutate(`PreSch Density` = `PreSch Count`/Area * 1000000)
```

## Explorotary Data Analysis (EDA)

Learn how to use appropriate ggplot2 functions to create functional and yet truthful statistical graphs for EDA purposes

First, plot a histogram to reveal the distribution of PreSch Density

```{r echo=TRUE, eval=TRUE}
hist(mpsz3414$`PreSch Density`)
```

```{r echo=TRUE, eval=TRUE}
ggplot(data=mpsz3414, 
       aes(x= as.numeric(`PreSch Density`)))+
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue") +
  labs(title = "Are pre-school even distributed in Singapore?",
       subtitle= "There are many planning sub-zones with a single pre-school, on the other hand, \nthere are two planning sub-zones with at least 20 pre-schools",
      x = "Pre-school density (per km sq)",
      y = "Frequency")
```

```{r echo=TRUE, eval=TRUE}
ggplot(data = mpsz3414,
       aes(x = as.numeric(`PreSch Density`), 
                          y= as.numeric(`PreSch Count`)))+
  geom_point()+
  labs(x = "Pre-school density (per km sq)",
       y = "Pre-school count")
  coord_cartesian(xlim=c(0,40),
                  ylim = c(0,40))
```

```{r echo=TRUE, eval=TRUE}

```

```{r echo=TRUE, eval=TRUE}

```

```{r echo=TRUE, eval=TRUE}

```

```{r echo=TRUE, eval=TRUE}

```

```{r echo=TRUE, eval=TRUE}

```

```{r echo=TRUE, eval=TRUE}

```

```{r echo=TRUE, eval=TRUE}

```
